
# Converting to Pearson Correlation

## From $t$-statistic

From a $t$ statistic calculated from a correlational test, we can calculate the correlation coefficient using the following formula:

$$
r = \sqrt{\frac{t^2}{t^2 + n-2}}
$$

Using the `t_to_r` function in the `effectsize` package we can convert $t$ to $r$.

```{r,echo=TRUE}
# Example:
# t = 4.14, n = 50

library(effectsize)

t <- 4.14
n <- 50

t_to_r(t = t, df = n-2)
```

## From Cohen's $d$

From a between groups Cohen's $d$ value ($d_p$), we can calculate the correlation coefficient from the following formula:

$$
r = \frac{d_p}{\sqrt{d_p^2+\frac{n_1+n_2-2}{n_1} + \frac{n_1+n_2-2}{n_2}}}
$$

Using the `d_to_r` function in the `effectsize` package we can convert $d_p$ to $r$.

```{r,echo=TRUE}
# Example:
# d = 0.60, n1 = 50, n2 = 70

d <- 0.60
n1 <- 50
n2 <- 70

d_to_r(d = d, n1 = n1, n2 = n2)
```

## From Odds-Ratio

The correlation coefficient from an odds ratio can be calculated with the following formula:

$$
r = \frac{\log(OR)\times\sqrt{3}}{\pi\sqrt{\frac{3\log(OR)^2}{\pi^2}+\frac{n_1+n_2-2}{n_1} + \frac{n_1+n_2-2}{n_2}}}
$$

Using the `oddsratio_to_r` function in the `effectsize` package we can convert $OR$ to $r$.

```{r,echo=TRUE}
# Example:
# OR = 2.21, n1 = 50, n2 = 70

OR <- 2.21
n1 <- 50
n2 <- 70

oddsratio_to_r(OR=OR, n1 = n1, n2 = n2)
```
