# Effect Sizes for ANOVAs

ANOVA (Analysis of Variance) is a statistical method used to compare means across multiple groups. It is mostly used when the outcome variable is continuous and the predictor variables are categorical. Commonly used effect size measures for ANOVAs / F-tests include: eta-squared ($\eta^2$), partial eta-squared ($\eta_p^2$), generalized eta-squared ($\eta^2_G$), omega-squared ($\omega^2$), partial omega-squared ($\omega$), generalized omega-squared ($\omega^2_G$), Cohen's $f$.

## Eta-Squared ($\eta^2$)

Eta-squared is the ratio between the between-group variance and the total variance. It describes the proportion of the total variability in the data that are accounted for by a particular factor. Therefore, it is a measure of *variance explained*. To calculate eta-squared ($\eta^2$) we need to first calculate the total sum of squares ($SS_{\text{total}}$) and the effect sum of squares ($SS_{\text{effect}}$),

$$
SS_{\text{total}} = \sum_{i=1}^n (y_i-\bar{y})^2
$$ 

Where $\bar{y}$ is the grand mean (i.e., the mean of all data points collapsed across groups). To calculate the sum of squares of the effect, we can take the predicted $y$ values ($\hat{y}_i$). In the case of categorical predictors, $\hat{y}_i$ is equal to the mean of the outcome *within* that individual's respective group. Therefore the sum of squares of the effect can be calculated using the following formula:

$$
SS_{\text{effect}} = \sum_{i=1}^n (\hat{y}_i-\bar{y})^2.
$$

Now we can calculate the eta-squared value,

$$
\eta^2 = \frac{SS_{\text{effect}}}{SS_{\text{total}}}
$$

The standard error of eta-square can be approximated from @olkin1995:

$$
SE_{\eta^2}=\sqrt{\frac{4\eta^2\left(1-\eta^2\right)^2\left(n+k-1\right)^2}{\left(n^2-1\right)\left(3+n\right)}}
$$

The sampling distribution for $\eta^2$ is asymmetric as all the values are bounded in the range, 0 to 1. The confidence interval surrounding $\eta^2$ will likewise be asymmetric so instead of calculating the confidence interval from the standard error, we can instead use a non-central F-distribution using the degrees of freedom between groups (e.g., for three groups: $df_b=k-1=3-1=2$) and the degrees of freedom within groups (e.g., for 100 subjects and three groups: $df_b=n-k=100-3=97$) to obtain the confidence intervals. Another option is to use bootstrapping procedure [i.e., resampling the observed data points to construct a sampling distribution around $\eta^2$, see @BootES] and then take the .025 and .975 quantiles of that distribution. The R code below will compute the proper confidence interval.

Where $n$ is the total sample size and $k$ is the number of predictors. In R, we can calculate $\eta^2$ from a one-way ANOVA using the penguin data set from the `palmerpenguins` data package. The `aov` function in base R allows the analyst to model an ANOVA with categorical predictors on the right side (species) of the `~` and the outcome on the left side (body mass of penguin). We can then use the `eta_squared` function in the `effectsize` package to calculate the point estimate and confidence intervals.

```{r,echo=TRUE}
# Example:
# group: species
# outcome: body mass

library(palmerpenguins)
library(effectsize)

# One-Way ANOVA
mdl1 <- aov(data = penguins,
           body_mass_g ~ species)

eta_squared(mdl1, 
            partial = FALSE,
            alternative = "two.sided")
```

The species of the penguin explains the majority of the variation in body mass showing an eta-squared value of $\eta^2$ = .67 \[.62, .71\]. Let us now do the same thing with a two-way ANOVA, using both `species` and `sex` as our categorical predictors.

```{r,echo=TRUE}
# Example:
# group: species and sex
# outcome: body mass

# Two-Way ANOVA
mdl2 <- aov(data = penguins,
           body_mass_g ~ species + sex)

eta_squared(mdl2, 
            partial = FALSE,
            alternative = "two.sided")
```

Notice that the $\eta^2$ does not change for species since the sum of squares is divided by the total sum of squares rather than the residual sum of squares (see partial eta squared). The example shows an eta-squared value for species of $\eta^2$ = .67 \[.62, .72\] and for sex $\eta^2$ = .17 \[.10, .24\].

## Partial Eta-Squared ($\eta^2_p$)

Partial eta-squared is the most commonly reported effect size measure for F-tests. It describes the proportion of variability associated with an effect when the variability associated with all other effects identified in the analysis has been removed from consideration (hence, it is "partial"). If you have access to an ANOVA table, the partial eta-squared for an effect is calculated as:

$$
\eta_p^2 = \frac{ SS_{\text{effect}}}{SS_{\text{effect}}+SS_{\text{error}}}
$$

There are two things to take note of here:

1.  In a one-way ANOVA (one categorical predictor), partial eta-squared and eta-squared are equivalent since $SS_{\text{total}} = SS_{\text{effect}}+SS_{\text{error}}$
2.  If there are multiple predictors, the denominator will only include the sum of squares of the effect of interest rather than the effect of all predictors (which is the case for the non-partial eta squared).

In R, let us compare the partial eta-squared values for a one-way ANOVA and a two-way ANOVA using the `eta_squared` function in the `effectsize` package.

```{r,echo=TRUE}
# Example:
# group: species
# outcome: body mass


# One-Way ANOVA
mdl1 <- aov(data = penguins,
           body_mass_g ~ species)

eta_squared(mdl1, 
            partial = TRUE,
            alternative = "two.sided") 
```

The species of the penguin explains the majority of the variation in body mass showing a partial eta-squared value of $\eta^2$ = $\eta^2_p$ = .67 \[.62, .71\]. Let us now do the same thing with a two-way ANOVA, using both `species` and `sex` as our categorical predictors.

```{r,echo=TRUE}
# Example:
# group: species and sex
# outcome: body mass

# Two-Way ANOVA
mdl2 <- aov(data = penguins,
           body_mass_g ~ species + sex)

eta_squared(mdl2, 
            partial = TRUE,
            alternative = "two.sided")
```

Once we run a two-way ANOVA, the eta-squared value for species begins to differ. The example shows a partial eta-squared value for species of $\eta^2_p$ = .81 \[.78, .84\] and for sex $\eta^2$ = .53 \[.46, .59\].

## Generalized Eta-Squared ($\eta^2_G$)

Generalized eta-squared was devised to allow effect size comparisons across studies with different designs, which eta-squared and partial eta-squared cannot help with (refer to for details). If you can (either you are confident that you calculated it right, or the statistical software that you use just happens to return this measure), report generalized eta-squared in addition to eta-squared or partial eta-squared. The biggest advantage of generalized eta-squared is that it facilitates meta-analysis, which is important for the accumulation of knowledge. To calculate generalized eta-squared, the denominator should be the sums of squares of all the non-manipulated variables (i.e., variance of purely individual differences in the outcome rather than individual differences in treatment effects). Note the formula will depend on the design of the study. In R, the `eta_squared` function in the `effectsize` package supports the calculation of generalized eta-squared by using the `generalized=TRUE` argument.

## Omega squared corrections ($\omega^2$, $\omega^2_p$)

Similar to Hedges' correction for small sample bias in standardized mean differences, $\eta^2$ is also biased. We can apply a correction to $\eta^2$ and obtain a relatively unbiased estimate of the population proportion of variance explained by the predictor. To calculate $\omega$, we need to calculate the within group mean squared errors:

$$
MS_{\text{within}} = \frac{1}{n}\sum_{i=1}^n (y_i-\hat{y}_i)^2.
$$ Where the predicted values of the outcome, $\hat{y}_i$, are the mean value for the individual's respective group.

$$
\omega^2 = \frac{SS_{\text{effect}}-(k-1)\times MS_{\text{within}}}{SS_{\text{total}} + MS_{\text{within}}}
$$

Where $k$ is the number of groups in the predictor (effect) variable. For partial omega-squared values, we need the mean squared error of effect and the residuals which can easily be calculated from their sum of squares:

$$
MS_{\text{effect}} = \frac{SS_{\text{effect}}}{n}
$$ $$
MS_{\text{error}} = \frac{SS_{\text{error}}}{n}
$$ Then to calculate the partial omega squared we can use the following formula:

$$
\omega_p^2 = \frac{(k-1)(MS_{\text{effect}} - MS_{\text{error}})}{(k-1)\times MS_{\text{effect}} + (n-k-1)\times MS_{\text{error}}}
$$

In R, we can use the `omega_squared` function in the `effectsize` package to calculate both $\omega^2$ and $\omega^2_p$. For the first example we will use a one-way ANOVA.

```{r,echo=TRUE}
# Example:
# group: species
# outcome: body mass

library(palmerpenguins)

# One-Way ANOVA
mdl1 <- aov(data = penguins,
           body_mass_g ~ species)

# omega-squared
omega_squared(mdl1, 
            partial = FALSE,
            alternative = "two.sided")

# partial omega-squared
omega_squared(mdl1, 
              partial = TRUE,
              alternative = "two.sided")
```

The species of the penguin explains the majority of the variation in body mass showing an omega-squared value of $\omega^2$ = .67 \[.61, .71\]. Note that the partial and non-partial omega squared values do not show a difference as expected in a one-way ANOVA. Let us now do the same thing with a two-way ANOVA, using both `species` and `sex` as our categorical predictors.

```{r,echo=TRUE}
# Example:
# group: species and sex
# outcome: body mass

# Two-Way ANOVA
mdl2 <- aov(data = penguins,
           body_mass_g ~ species + sex)

# omega-squared
omega_squared(mdl2, 
            partial = FALSE,
            alternative = "two.sided")

# partial omega-squared
omega_squared(mdl2, 
              partial = TRUE,
              alternative = "two.sided")
```

Once we run a two-way ANOVA, the eta-squared value for species  diverge. The example shows a partial eta-squared value for species of $\omega^2_p$ = .81 \[.78, .84\] and for sex $\omega^2$ = .53 \[.46, .58\].

### Cohen's $f$

Cohen's $f$ is defined as the ratio of the standard deviations of the group means and the common standard deviation within each of the groups (note that ANOVA assumes equal variances among groups). Cohen's $f$ is the effect size measure asked for by G\*Power for power analysis for F-tests. This can be calculated easily from the eta-squared value,

$$
f = \sqrt{\frac{\eta^2}{1-\eta^2}}
$$

or by the $\omega^2$ value,

$$
f = \sqrt{\frac{\omega^2}{1-\omega^2}}
$$

Cohen's $f$ can be interpreted as "the average Cohen's $d$ (i.e., standardized mean difference) between groups". Note that there is no directionality to this effect size ($f$ is always greater than zero), therefore two studies showing the same $f$ with the same groups, can have very different patterns of group mean differences. Note that Cohen's $f$ is also often reported as $f^2$. The confidence intervals for Cohen's $f$ can be computed from the upper bounds and lower bounds of the confidence intervals from eta-square or omega-square using the formulas to calculate $f$ (e.g., for the upper bound $f_{UP} = \sqrt{\frac{\eta^2_{UP}}{1-\eta^2_{UP}}}$).

In R, we can use the `cohens_f` function in the `effectsize` package to calculate Cohen's $f$. We will again use example data from the `palmerpenguins` package.

```{r,echo=TRUE}
# Example:
# group: species
# outcome: body mass

# ANOVA
mdl <- aov(data = penguins,
           body_mass_g ~ species)   

cohens_f(mdl,alternative = "two.sided")

```

In the example above, the difference in body mass between the three penguin species was very large showing a Cohen's $f$ of 1.42 \[1.27, 1.57\].

## Reporting ANOVA results

For ANOVAs/F-tests, you will always need to report two kinds of effects: the omnibus effect of the factor(s) and the effect of planned contrasts or post hoc comparisons.

For instance, imagine that you are comparing three groups/conditions with a one-way ANOVA. The ANOVA will first return an F-statistic, the degrees of freedom, and the associated p-value. Here, you need to calculate the size of this omnibus factor effect in eta-squared, partial eta-squared, or generalized eta-squared. Suppose the omnibus effect is significant. You now know that there is at least one group that differs from the others. You want to know which group(s) differ from the others, and how much they differ. Therefore, you conduct post hoc comparisons on these groups. Because post hoc comparisons compare each group with the others in pairs, you will get a *t*-statistic and p-value for each comparison. For this, you need to calculate and report Cohen's $d$ or Hedges' $g$.

Imagine that you have two independent variables or factors, and you conduct a two-by-two factorial ANOVA. The first thing to do then is look at the interaction. If the interaction is significant, you again report the associated omnibus effect size measures, and proceed to analyze the simple effects. Depending on your research question, you compare the levels of one IV on each level of the other IV. You will report d or g for these simple effects. If the interaction is not significant, you look at the main effects and report the associated omnibus effect. You then proceed to analyze the main effect by comparing the levels of one IV while collapsing/aggregating the levels of the other IV. You will report $d$ or $g$ for these pairwise comparisons.

Note that lower-order effects are not directly interpretable if higher-order effects are significant. If you have a significant interaction in a two-way ANOVA, you cannot interpret the main effects directly. If you have a significant three-way interaction in a three-way ANOVA, you cannot interpret the main effects or the two-way interactions directly, regardless of whether they are significant or not.

In R, we can use the `summary` function to display the anova table. We can also append the table to include, for example, partial omega squared values and their respective confidence intervals

```{r,echo=TRUE}
# ANOVA mdl
mdl <- aov(data = penguins,
           body_mass_g ~ species + sex)   

# calculate partial omega-squared values
omega_values <- omega_squared(mdl, alternative = "two.sided")

# create table of partial omega-squared values
omega_table <- data.frame(omega_sq = MOTE::apa(c(omega_values$Omega2_partial,NA)),
                     omega_low = MOTE::apa(c(omega_values$CI_low,NA)),
                     omega_high = MOTE::apa(c(omega_values$CI_high,NA)))

# append omega values to summary of anova table
cbind(summary(mdl)[[1]],
      omega_table)
```

